### 2018-07-09 latex/introduction.tex
further editing of introduction and results first paragraph completed

Due to the constant release of Nam through NAD-consuming signalling reactions, the NAD salvage pathway using Nam as precursor is the most important NAD synthesis pathway. If Nam would not be constantly recycled into NAD, humans would require a much higher daily vitamin B3 intake than the 16\,mg that are the current daily recommendation \cite{CommissionofEuropeanCommunities2008}. Two principal pathways exist that recycle Nam. Firstly, vertebrates use a direct two step pathway staring with the convertion of Nam into the mononucleotide NMN by the Nam phosphoribosyltransferase (NamPRT) using phosphoribosyl pyrophosphate (PRPP) as co-substrate. The nearly complete recycling of Nam by NamPRT is achieved by an extraordinary high subtrate affinity to Nam, the $K_{M}$ being in the low nanomolar range \cite{Burgos2008}. This appears to be mediated by an ATP-dependent phosphorylation of a histidine residue in the catalytic core\todo{PMID:19666527}. Despite the importance of its salvage, Nam can also be marked for excretion by methylation. The presence of nicotinamide N-methyltransferase (NNMT) in vertebrates \cite{Gossmann2012FEBS} is among the most enigmatic and counterintuitive features of NAD metabolism. While NamPRT is seemingly optimised to recycle even the faintest amounts of Nam back into NAD synthesis, NNMT seems to have no metabolic function other than to remove Nam from NAD metabolism. Although it has been suggested that it potentially acts as methylation sink \cite{NNMTreview2018} . Methyl-Nam produced by NNMT is in mammals excreted with the urine, thus removing Nam from recycling.

Secondly, in many prokayotes as well as in plants and fungi a four steps starting with the deamidation of Nam to nicotinic acid (NA) by NADA is used. (\ref{fig:pathway_overview}). The three enzymes that act after NADA belong to the Preiss-Handler pathway that also exists in vertebrates. The Preiss-Handler pathway first converts  NA into the corresponding mononucleotide (NAMN), in a reactions performed by the NAD specific phosphoribosyl transferase NAPRT. The convertion of the mononucleotides  NMN and NAMN into their dinucleotides NAD and NaAD is performed by the same enzyme, the  Nam/NA adenylytransferases (NMNATs) that is essential in all Organisms\cite{Gossmann2011}. The recycling pathway via NA finally requires re-amidation of the dinucleotide, which is catalysed by NAD synthase. This reaction includes an enzyme adenylation step that consumes ATP. Therefore the Nam recycling by NADA is energetically less efficient under normal conditions.

>>>>

Due to the constant release of Nam through NAD-consuming signalling reactions, the NAD salvage pathway using Nam as precursor is the most important NAD synthesis pathway. If Nam would not be constantly recycled into NAD, humans would require a much higher daily vitamin B3 intake than the 16\,mg that are the current daily recommendation \cite{CommissionofEuropeanCommunities2008}. Two principal pathways exist that recycle Nam. Firstly, vertebrates use a direct two step pathway staring with the convertion of Nam into the mononucleotide NMN by the Nam phosphoribosyltransferase (NamPRT) using phosphoribosyl pyrophosphate (PRPP) as co-substrate. The nearly complete recycling of Nam by NamPRT is achieved by an extraordinary high subtrate affinity to Nam, the $K_{M}$ being in the low nanomolar range \cite{Burgos2008}. This appears to be mediated by an ATP-dependent phosphorylation of a histidine residue in the catalytic core\todo{PMID:19666527}. Despite the importance of its salvage, Nam can also be marked for excretion by methylation. The presence of nicotinamide N-methyltransferase (NNMT) in vertebrates \cite{Gossmann2012FEBS} is among the most enigmatic and counterintuitive features of NAD metabolism. While NamPRT is seemingly optimised to recycle even the faintest amounts of Nam back into NAD synthesis, NNMT seems to have no metabolic function other than to remove Nam from NAD metabolism. Although it has been suggested that the process potentially acts as a metabolic methylation sink \cite{PMID:28291578} .

Secondly, in many prokayotes as well as in plants and fungi a four steps starting with the deamidation of Nam to nicotinic acid (NA) by NADA is used. (\ref{fig:pathway_overview}). The three enzymes that act after NADA belong to the Preiss-Handler pathway that also exists in vertebrates. The Preiss-Handler pathway first converts  NA into the corresponding mononucleotide (NAMN), in a reactions performed by the NAD specific phosphoribosyl transferase NAPRT. The convertion of the mononucleotides  NMN and NAMN into their dinucleotides NAD and NaAD is performed by the same enzyme, the  Nam/NA adenylytransferases (NMNATs) that is essential in all Organisms\cite{PMID:21729004}. The recycling pathway via NA finally requires re-amidation of the dinucleotide, which is catalysed by NAD synthase. This reaction includes an enzyme adenylation step that consumes ATP. Therefore the Nam recycling by NADA is energetically less efficient under normal conditions.

### 2018-07-09 latex/results.tex
further editing of introduction and results first paragraph completed

\subsection{Paradoxical evolutionary correlation between NAD-dependent signalling and precursor metabolism}
As the Preiss-Handler pathway  is  present in virtually all organisms analysed previously\cite{Gossmann2012FEBS} and as based on metabolic pathway analysis we have previously shown tha NMNAT is essential for NAD dependent biosynthesis independent of the biosynthetic route or precursor use \cite{Gossmann2011} we here focused on the phylogenetic analysis of the two enzymes that initiate the two different NAD-salvage pathways NADA and NamPRT as well as on the NAM methyltransferase NNMT as the distribution of these enzymes seems to differ substantially between different organisms. As can be seen from Fig. (fig.~\ref{fig:phylo_distribution}A, bacteria, fungi and plants predominantly encode NADA and only a few of them harbour NamPRT.  In contrast, Metazoa predominantly have NamPRT together with NNMT. Analysing the phylogeny distribution in Metazoa in more detail revealed that not only is NamPRT replaced by NADA, but in most organisms, especially in Deuterostomes, NamPRT is found together with NNMT (fig.~\ref{fig:phylo_distribution}B). NNMT seems to have arisen \textit{de~novo} in the common ancestor of Ecdysozoa and Lophotrochozoa. We were unable to find any \textit{NNMT} gene with considerable \todo{define considerable} similarity in fungi or plants.

Nematodes are the only organisms, where we observed any combination of NADA, NamPRT, and NNMT. In Deuterostomes, the only large clade that possesses only NamPRT and seems to have lost NNMT again are Sauropsida and among them especially birds. The reason why about half of the sequenced bird genomes do not seem to encode NNMT remains unclear, as the distribution is quite scattered (suppl. fig.~S2). The lack of NNMT might be related to the excretion system, as the product of NNMT, methyl-Nam, is in mammals excreted with the urine. There are some Metazoan species for which we could not find NamPRT or NADA, while NNMT was present. We assume that this is due to incomplete genomes in the database, as the distribution of such species is scarce and widely scattered.

@@

Taken together, we found that NADA is lost in vertebrates, but strongly preserved in most other organisms, despite the higher energetic requirement of that pathway. Moreover, the selection for having both NamPRT and NNMT coincides with an increased diversification of NAD-dependent signalling. This observation seems counterintuitive. One would expect that increased NAD-dependent signalling should be accompanied by an increase of substrate availability for NAD biosynthesis. However, NNMT irreversibly removes Nam from salvage back to NAD.

>>>>

\subsection{Paradoxical evolutionary correlation between NAD-dependent signalling and precursor metabolism}
We have  previously shown  that NMNAT is essential for NAD dependent biosynthesis independent of the biosynthetic route or precursor \cite{PMID:21729004}. In addition, the Preiss-Handler pathway was found  present in virtually all organisms included in an earlier phylogenetic analysis\cite{Gossmann2012FEBS}. Thus, we here focused on the phylogenetic analysis of the two enzymes that initiate the two different NAD-salvage pathways, NADA and NamPRT. These show a scattered distribution in bacteria \todo{MathiasB: please find and insert citation} and have been shown to occur concomitantly in some marine invertebrates \cite{Gossmann2012FEBS}. In addition, we included  NAM methyltransferase NNMT into our analysis, as its function and origin is unclear. 
As can be seen from Fig. (fig.~\ref{fig:phylo_distribution}A, bacteria, fungi and plants predominantly encode NADA and only a few of them harbour NamPRT.  In contrast, Metazoa predominantly have NamPRT together with NNMT. Analysing the phylogeny distribution in Metazoa in more detail revealed that not only is NamPRT dominating and NADA appears to be lost in vertebrates, in most organisms NamPRT is found together with NNMT (fig.~\ref{fig:phylo_distribution}B). NNMT seems to have arisen \textit{de~novo} in the common ancestor of Ecdysozoa and Lophotrochozoa. As we were unable to find any \textit{NNMT} gene with an e-value below .... \todo{Mathias B. include e-value}  in fungi or plants. The gene with the highest sequence similarity in S. cerevisiae is  ..... with proven function .....\todo{Mathias B: include e-value}

Nematodes are the only organisms, where we observed a concomitant presence of NADA and NNMT. In Deuterostomes, the only large clade that possesses only NamPRT and seems to have lost NNMT are Sauropsida, and among them especially birds. The reason why about half of the sequenced bird genomes do not seem to encode for  NNMT,  remains unclear. The distribution of NNMT in birds is quite scattered (suppl. fig.~S2). not corresponding to current species trees \cite{http://dx.doi.org/10.1038/nature15697} . The lack of NNMT might be related to the excretion system, as the product of NNMT, methyl-Nam, is in mammals excreted with the urine. There few Metazoan species for which we could not find NamPRT or NADA, while NNMT was present. We assume that this is due to incomplete genomes in the database, as the distribution of such species is scarce and widely scattered.

@@

Taken together, we found that NADA is lost in vertebrates, but strongly preserved in most other organisms, despite the higher energetic requirement of that pathway. Moreover, the selection for having both NamPRT and NNMT coincides with an increased diversification of NAD-dependent signalling. This observation seems counterintuitive, as one would expect that increased NAD-dependent signalling should be accompanied by an increase of substrate availability for NAD biosynthesis. However, NNMT irreversibly removes Nam from salvage and thus is assumed to potentially decrease the cellular NAD-concentration.
